Assessing primitives performance\\ on multi-stage execution

### 2018-05-08 paper.tex
reading up where it makes sense

High-level object-oriented programming languages are often implemented on top of a virtual machine (VM). As VMs has become more popular, different techniques have been set-up using Just-In-Time (JIT) compilation to improve the overall performance of the runtime such as method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.

On top of its virtual machine, the client language can use a set of \emph{primitives}~\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation. Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method. We distinguish two kinds of primitives:
\begin{itemize}
	\item \emph{Essential primitives} cannot be performed in any other way. A high-level object oriented language without primitives can move values from one variable to another, but cannot add two integers together. Many arithmetic and comparison operations between numbers are primitives. Some primitives allow to communicate with I/O devices such as the disk, the display and the keyboard. 
	\item \emph{Optional primitives} exist only to make the system run faster. They can be implemented in the client language directly, making its implementation as a primitive optional, but they are implemented in the VM to improve performance.
\end{itemize}

The implementation of an optional primitive in the VM rather than the client language improves performance but also often increases the implementation engineering cost. For example, accessing objects in the VM usually requires some understanding of the memory layout or implementation details, which are not needed in the client language.

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
	\item The performance of the code not yet optimized is much worse than the peak performance code,
	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

VMs are traditionally implemented in a low-level language such as C or C++. To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler at runtime the frequently used code snippets for further executions to be performed quicker. This leads to annoying concerns: for example, let's say we implement a primitive in C. Frequently used portion of code calling that primitive are going to be compiled to native code by the JIT. This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers. Switching between both can waste up to around a dozen native instructions. If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.

In this paper, we take the example of a string comparison optional primitive. The primitive compares two strings and answers if one string is greater, equal or lower than the other string. The exact specifications are detailed in Section \ref{subsec:primSpec}. We implemented a first version in Smalltalk. We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.

@@

Our evaluation is based on the Cog VM a Smalltalk virtual machine. The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code. Historically, the VM implemented a 32-bits adaptation of the Smalltalk-80 specifications \cite{blueBook}, but the Cog VM is now the default VM for multiple programming languages such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}.

Section \ref{sec:VMCompil} describes briefly the implementation languages and compilation process of the Cog VM. Section \ref{sec:VMexec} explains briefly the hybrid execution model with an interpreter and a JIT. In Section \ref{sec:primExec}, we detail how primitives are executed. Section \ref{subsec:primSpec} discusses the specifications of the string comparison primitive used for our evaluation.

@@

Most of the Cog VM code base is written in Slang, a subset of Smalltalk. Slang is compiled to C and then to native code through standard C compilers. The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang. The two main purposes of using slang code over plain C are:
\begin{itemize}
	\item to specify with annotations what function needs to be inlined or duplicated with constant parameters, and
	\item to be able to simulate VM execution, executing the Slang code as Smalltalk code on top of a compiled VM for debugging purposes.
\end{itemize}

@@

The executable is generated in two steps as shown on Figure \ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}. The first step is to generate the two C files representing the whole execution engine written in Slang using the Slang-to-C compiler. During the second step, a C compiler is called to compile the execution engine and the platform-specific code written directly in C to the executable VM.

@@

The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime. During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure \ref{fig:TwoStacks}. This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack. When executing the native code generated by JIT, only the client language stack is used. This means for example that the frame pointer register in the processor refers to a stack frame on the client stack and that a processor push instruction pushes a value on the client stack.

@@

To solve this performance bottleneck, the Cog VM allows to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end. Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code. This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance. Primitives redefined in such way can be entirely or partially redefined. In the later case, only the common cases are re-implemented on top of the JIT back-end, and an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.

@@

\paragraph{String representations.} Strings are represented in all programming languages on top of the Cog VM in two possible forms, ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. WideStrings encode each character of the string in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.

\paragraph{Original specification.}  The original primitive took three operands. The two first operands were the two ByteStrings to compare and the third operand is the order table, a ByteArray encoding the order of characters. The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i). However, in some case, the program may do case insensitive ByteStrings comparison. In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value). To compare the two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString. If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns. Based on the three parameters, the primitive answered 3,2 or 1 depending if the first ByteString was greater, equal or smaller than the second one.

\paragraph{Specification issues.} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
\begin{itemize}
	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1,2 or 3. It looks like a minor difference, but it is quite important for performance. With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal. With 1,2 or 3, the primitive has to end with multiple branches to know what number to return. In native code, a subtraction is on common architecture quicker to perform than one or multiple branches. In addition, the primitive code is easier to write with the subtraction when writing directly the assembly instructions and we did exactly that when writing the primitive on top of the JIT's back-end.
	\item \emph{Order:} Having the order as the third operands is convenient in some cases (case insensitive comparison), but is just overhead in the common case (checking for string equality).
\end{itemize}

@@

 This section first describes the different ways of implementing a primitive. We then explain how each of these version are executed by the VM and the pros and cons of each and every.

>>>>

High-level object-oriented programming languages are often implemented on top of a virtual machine (VM). As VMs has become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the runtime overall performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.

On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation. Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method. We distinguish two kinds of primitives:
\begin{itemize}
	\item \emph{Essential primitives} cannot be performed in any other way. A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together. Many arithmetic and comparison operations between numbers are primitives. Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard. 
	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}. They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
\end{itemize}

The implementation of an optional primitive in the VM rather than the client language improves performance but also often increases the implementation engineering cost. For example, accessing objects in the VM usually requires understanding of the memory layout or implementation details, which are not needed in the client language.

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. \sd{sentence was too long: please reread my version:}Their goal was to use a JIT performing speculative optimizations of optional primitives implemented in the client language. Ideally we wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
	\item The performance of the code not yet optimized is much worse than the peak performance code,
	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

VMs are traditionally implemented in a low-level language such as C or C++. To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler \sd{optimizes} at runtime the frequently used code snippets \sd{rm: for further executions to be performed quicker}.

This leads to annoying concerns: for example, let's say we implement a primitive in C. Frequently used portion of code calling that primitive are going to be compiled to native code by the JIT. This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers. Switching between both usually wastes up to around a dozen native instructions. If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.

\sd{the link is missing: In this article we want to measure the complexities and gain of this different implementations}

In this paper, we take the example of a string comparison optional primitive. The primitive compares two strings and answers if one string is greater, equal or lower than the other string. The exact specifications are detailed in Section \ref{subsec:primSpec}. We implemented a first version in Pharo/Smalltalk. We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.

@@

Our evaluation is based on the Cog VM a Smalltalk virtual machine. The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and \sd{has} have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code. Historically, the VM implemented a 32-bits adaptation of the Smalltalk-80 specifications \cite{blueBook}, but the Cog VM is now the default VM for multiple programming languages such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}.

Section \ref{sec:VMCompil} describes briefly the implementation languages and compilation process of the Cog VM. Section \ref{sec:VMexec} explains briefly the hybrid execution model with an interpreter and a JIT. In Section \ref{sec:primExec}, we detail how primitives are executed. Section \ref{subsec:primSpec} discusses the string comparison primitive specifications  used for our evaluation.

@@

Most of the Cog VM code base is written in Slang, a subset of Smalltalk. Slang is compiled to C and then to native code through standard C compilers. The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang. The two main purposes of using Slang code over plain C are:
\begin{itemize}
	\item to specify with annotations what function needs to be inlined or duplicated with constant parameters, and
	\item to be able to simulate VM execution, executing the Slang code as Smalltalk code on top of a compiled VM for debugging purposes.
\end{itemize}

@@

The executable is generated in two steps as shown on Figure~\ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}. The first step is to generate the two C files representing the whole execution engine written in Slang using the Slang-to-C compiler. During the second step, a C compiler is called to compile the execution engine and the platform-specific code written directly in C to the executable VM.

@@

The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime. During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure~\ref{fig:TwoStacks}. This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack. When executing the native code generated by JIT, only the client language stack is used. This means for example that the frame pointer register in the processor refers to a stack frame on the client stack and that a processor push instruction pushes a value on the client stack.

@@

To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end. Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code. This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance. Primitives redefined in such way can be entirely or partially redefined. In the later case, only the common cases are re-implemented on top of the JIT back-end, and an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.

\paragraph{Original specification.}  The original primitive took three operands. The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters. The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i). However, in some case, the program may do case insensitive ByteStrings comparison. In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value). To compare two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString. If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns. Based on the three parameters, the primitive answered 3, 2, or 1 depending if the first ByteString was greater, equal or smaller than the second one.

\paragraph{Specification issues.} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
\begin{itemize}
	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1, 2 or 3. It looks like a minor difference, but it is important for performance. With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal. With 1, 2, or 3, the primitive has to end with multiple branches to know what number to return. In native code, a subtraction is on common architecture quicker to perform than one or multiple branches. In addition, the primitive code is easier to write with the subtraction when writing directly the assembly instructions and we did exactly that when writing the primitive on top of the JIT's back-end.
	\item \emph{Order:} Having the order as the third operands is convenient in some cases (case insensitive comparison), but is just overhead in the common case (checking for string equality).
\end{itemize}

@@

 This section first describes the different ways of implementing a primitive. We then explain how each of these versions are executed by the VM and the pros and cons of each and every.
